Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach

### 2024-11-21 paper/paper.qmd
update paper with model and data part

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.31 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Our Results show that the overturning of Roe v. Wade and the implementation of abortion bans are linked to a rise in infant mortality rates in states where abortion became illegal. Through a Difference-in-Differences (DID) approach, we estimate that states enforcing abortion bans after the injunction saw an increase of 0.31 deaths per 1,000 live births compared to states where abortion remained legal. Furthermore, we find that maternal age and race are key determinants of infant mortality, with older maternal age groups associated with lower mortality rates and Black mothers experiencing significantly higher rates. These results emphasize the public health consequences of restrictive abortion laws and the urgent need for policies that mitigate these disparities.

@@

@@

ggplot(analysis_data, aes(x = death_rate)) +
  geom_histogram(binwidth = 1, fill = "brown", color = "black") +
  labs(
    x = "Infant Death Rate (per 1,000 live births)", y = "Count",
    caption = "Data source: CDC Wonder."
  ) +theme_classic()
```
@fig-deathrate illustrates the distribution of infant death rates (per 1,000 live births) based on the available data. The majority of infant death rates cluster around lower values, with a peak observed between 5 and 10 deaths per 1,000 infants. As the death rate increases, the frequency of occurrences decreases, indicating that higher infant death rates are relatively rare. There are few instances where the rate exceeds 20 deaths per 1,000 infants, suggesting these are outliers. Overall, the distribution is right-skewed, highlighting the predominance of lower infant death rates in the dataset.

@@

```{r}
#|include: false
library(usmap)
library(ggplot2)
library(dplyr)

@@

```{r}
#|include: false
us_map <- map_data("state")
map_data <- us_map|>
   left_join(abortion_status, c("region"="state"))|>filter(!is.na(abortion_illegal))
```

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) 
```
@fig-map illustrates the legality of abortion across U.S. states. Based on @reproductiveright, as of November 2024, abortion has been illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin following the June 2022 decision.
```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() 
```
@fig-rateandstate illustrates trends in infant death rates (per 1,000 live births) over time in states where abortion is illegal versus legal. In states with abortion bans, infant mortality shows a slight increase after June 2022, marked by the dashed line representing the overturn of Roe v. Wade. In contrast, states where abortion remains legal exhibit relatively stable trends. This suggests a potential link between abortion bans and rising infant mortality rates, warranting further investigation.

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with “Illegal Abortion and After Injunction” (denoted as ‘1’) have a higher average infant death rate compared to states with “Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction” (denoted as ‘0’).
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`'0': Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction, '1': Illegal Abortion and After Injunction`=illegal_after_injunction)

# Display the result as a table
kable(analysis_data_state)|>footnote("Data Source: CDC Wonder")
```
@tbl-interact highlights average infant death rates (per 1,000 live births) based on abortion legality and injunction timing. States categorized as ‘1’, where abortion became illegal after the injunction, show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’ states, where abortion remained legal or abortion used to be legal before the injunction. This suggests that restrictive abortion policies and their enforcement may contribute to higher infant mortality, potentially due to increased unintended or high-risk pregnancies, reduced access to maternal healthcare, and exacerbated social or economic disparities.

@@

@@

ggplot() +
  # First layer: Fill with death_rate
  geom_polygon(
    data = death_full,
    aes(x = long, y = lat, group = group, fill = death_rate),
    color = "black"
  ) +
  scale_fill_viridis_c(
    name = "Infant Death Rate\n(per 1,000 live births)",
    option = "C",
    na.value = "grey50",
    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
  ) +
  theme_void() +
  labs(
    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
  ) 
```
@fig-choromap is a choropleth map visualizes the average infant death rate per 1,000 live births across U.S. states, with data categorized by color intensity. States with higher infant death rates are shown in lighter shades (yellow), while states with lower rates are represented in darker shades (purple). States with missing data are indicated in grey. The map highlights disparities in infant mortality rates, with some states, particularly in the southeastern U.S., exhibiting notably higher rates compared to others. States like Alabama and Mississippi show the highest rates, while states such as California and New York exhibit lower rates.

@@

@tbl-motherage displays the average infant death rate (per 1,000 live births) by the age group of mothers. It shows a clear trend: younger mothers, particularly those aged 15–19 and 20–24, have higher infant death rates (9.0 and 10.0, respectively), while older mothers in the 25–29 and 30–34 age groups have lower rates (7.4 and 6.6, respectively). This suggests that maternal age may be a factor influencing infant mortality, with younger mothers experiencing higher risks.
```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

### Mother's Single Race
The self-identified race of the mother, recorded as a single race category. In this dataset, the available categories are: Asian, Black or African American, and White. This classification facilitates the analysis of maternal and infant health outcomes across these racial groups.
```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

# Appendix {-}

@@

@@

```{r}
analysis_data|>
  ggplot(aes(x = after_injunction, y = death_rate, color = as.factor(abortion_illegal))) +
  geom_point(alpha = 0.2) +
  geom_line(aes(group = state)) +
  theme_minimal() +
  labs(x = "Time period", y = "Death Rate", color = "Person got a new racket")  +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

>>>>

---
title: "Examining the impact Overturn of Roe v. Wade: Banning of Abortion on Infant Mortality Rates in the United States Using a Difference-in-Differences Approach"
subtitle: "The infant death rate increased by 0.285 deaths per 1,000 live births in states that became abortion-banned, compared to states where abortion remained legal."
author: 
  - Diana Shen
thanks: "Code and data are available at: [https://github.com/DianaShen1224/Relationship-between-infant-mortality-rate-and-prohibited-abortion)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Our Results show that the overturning of Roe v. Wade and the
implementation of abortion bans are linked to a rise in infant mortality
rates in states where abortion became illegal. Through a
Difference-in-Differences (DID) approach, we estimate that states
enforcing abortion bans after the injunction saw an increase of 0.285
deaths per 1,000 live births compared to states where abortion remained
legal. Furthermore, we find that maternal age and race are key
determinants of infant mortality, with older maternal age groups
associated with lower mortality rates and Black mothers experiencing
significantly higher rates. These results emphasize the public health
consequences of restrictive abortion laws and the urgent need for
policies that mitigate these disparities.

@@

The process of translating real-world events into our dataset requires a
systematic approach to measurement and data gathering. In this research,
we analyze infant mortality rates in the United States to understand the
impact of abortion policies and demographic factors. The CDC’s Linked
Birth/Infant Death Records database provides comprehensive data,
including maternal and infant characteristics, to support this study.
Variables such as maternal age, race, education, birth weight,
gestational age, and prenatal care are included to ensure a detailed
examination of factors influencing infant health outcomes.

The data, sourced from birth and death certificates, undergoes a
thorough cleaning and validation process to address inconsistencies and
missing information, ensuring accuracy. Infant mortality rates are
calculated per 1,000 live births, with additional indicators such as
reliability labels for rates based on fewer than 20 deaths. To further
enhance data accuracy, weights are applied to adjust for unlinked infant
death records, ensuring that all deaths—linked and unlinked—are
proportionally represented. Weights are calculated for each state and
infant age-at-death cohort (\<7 days, 7-27 days, and 28 days to 1 year)
using the formula:
$\text{Weight} = \frac{\text{Linked deaths} + \text{Unlinked deaths}}{\text{Linked deaths}}$
This adjustment means areas with complete linkages have a weight of 1.0,
while areas with incomplete linkages are adjusted proportionally.
Additionally, imputation techniques are employed to handle missing
birthweight data, assigning values based on gestational period, race,
sex, and plurality, further reducing potential bias in the analysis.

Once cleaned, the data is stratified by maternal and infant attributes,
providing insights into trends and disparities across demographic
groups. Statistical models are employed to estimate associations between
state-level abortion policies and infant mortality, controlling for
confounders such as maternal characteristics. This structured approach
transforms complex datasets into actionable insights, shedding light on
the health implications of varying abortion laws across the U.S. By
leveraging comprehensive datasets and robust statistical techniques,
this study offers a nuanced understanding of the intersection between
policy, demographics, and infant health outcomes, informing public
health strategies and legislative decisions.

@@

ggplot(analysis_data, aes(x = death_rate)) +
  geom_histogram(binwidth = 1, fill = "brown", color = "black") +
  labs(
    x = "Infant Death Rate (per 1,000 live births)", y = "Count",
    caption = "Data source: CDC Wonder."
  ) +theme_classic()
```

@fig-deathrate illustrates the distribution of infant death rates (per
1,000 live births) based on the available data. The majority of infant
death rates cluster around lower values, with a peak observed between 5
and 10 deaths per 1,000 infants. As the death rate increases, the
frequency of occurrences decreases, indicating that higher infant death
rates are relatively rare. There are few instances where the rate
exceeds 20 deaths per 1,000 infants, suggesting these are outliers.
Overall, the distribution is right-skewed, highlighting the predominance
of lower infant death rates in the dataset.

@@

library(usmap)
library(ggplot2)
library(dplyr)

@@

```{r}
#| include: false
#| warning: false
#| message: false
us_map <- map_data("state")
map_data <- us_map|>
   left_join(abortion_status, c("region"="state"))|>filter(!is.na(abortion_illegal))
```

```{r}
#| label: fig-map
#| fig-cap: Abortion Legality by State. As of November 2024, abortion is illegal in Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi, Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West Virginia, and Wisconsin.
#| echo: false
ggplot() +
  geom_polygon(data = map_data, aes(fill = abortion_illegal, x = long, y = lat, group = group),
    color = "black") +
  scale_fill_manual(
    values = c("0" = "blue", "1" = "red"),
    labels = c("Legal", "Illegal"),
    name = "Abortion Legal Or Not"
  )+
  theme_void() +
  labs(
    caption = "Source: Center for Reproductive Rights\nDate as of Nov.19 2024"
  ) +
  theme(legend.position = "bottom")
```

@fig-map illustrates the legality of abortion across U.S. states. Based
on @reproductiveright, as of November 2024, abortion has been illegal in
Alabama, Arkansas, Idaho, Indiana, Kentucky, Louisiana, Mississippi,
Missouri, North Dakota, Oklahoma, South Dakota, Tennessee, Texas, West
Virginia, and Wisconsin following the June 2022 decision.

```{r}
#| label: fig-rateandstate
#| fig-cap: Infant death rates (per 1,000 live births) over time, with a slight increase after June 2022 in states where abortion is illegal, while rates remain stable in states where abortion is legal.
#| echo: false
#| message: false
analysis_data|>
  mutate(illegal = if_else( abortion_illegal== 1, "Illegal State", "Legal State"))|>
  ggplot(aes(x = date, y = death_rate,colour = illegal)) +
  geom_point(alpha = 0.5)+
  geom_smooth() +
  labs(x = "Date of Infant Death", y = "Infant Death Rate (per 1,000 live births)",color="State is Illegal for Abortion Or Not",caption = "Data source: CDC Wonder.\nIn June 2022, the U.S. Supreme Court overturned Roe v. Wade.") +
  facet_wrap(vars(illegal), nrow = 2)+
  geom_vline(xintercept = as.Date("2022-06-01"), linetype = "dashed")  +
  scale_x_date(
    limits = c(as.Date("2021-01-01"), as.Date("2022-12-01")), 
    date_breaks = "3 months",  
    date_labels = "%Y-%m" 
  )+
  scale_color_brewer(palette = "Set1")+
  theme_classic() +
  theme(legend.position = "bottom")
```

@fig-rateandstate illustrates trends in infant death rates (per 1,000
live births) over time in states where abortion is illegal versus legal.
In states with abortion bans, infant mortality shows a slight increase
after June 2022, marked by the dashed line representing the overturn of
Roe v. Wade. In contrast, states where abortion remains legal exhibit
relatively stable trends. This suggests a potential link between
abortion bans and rising infant mortality rates, warranting further
investigation.

```{r}
#| label: tbl-interact
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) by Interaction Between Abortion Legality and Injunction Timing. States with "Illegal Abortion and After Injunction" (denoted as '1') have a higher average infant death rate compared to states with "Legal Abortion and Before/After Injunction or Illegal Abortion and Before Injunction" (denoted as '0').
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>mutate(illegal_after_injunction=as.numeric(as.character(abortion_illegal))*as.numeric(as.character(after_injunction)))|>
  group_by(illegal_after_injunction)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1))

kable(analysis_data_state, align = "c") |> 
  footnote(
    "Data Source: CDC Wonder"
  )
```

@tbl-interact highlights average infant death rates (per 1,000 live
births) based on abortion legality and injunction timing. States
categorized as ‘1’, where abortion became illegal after the injunction,
show a higher average infant death rate of 8.3 compared to 7.9 in ‘0’
states, where abortion remained legal or abortion used to be legal
before the injunction. This suggests that restrictive abortion policies
and their enforcement may contribute to higher infant mortality,
potentially due to increased unintended or high-risk pregnancies,
reduced access to maternal healthcare, and exacerbated social or
economic disparities.

@@

The Year of Death variable represents the calendar year in which the
infant death occurred, specifically focusing on the years 2021 and 2022
in this study. This variable is essential for analyzing temporal trends
and identifying patterns or shifts in infant mortality rates over this
critical period. By controlling the year of death, we can closely
examine the impact of significant policy changes, such as the
overturning of Roe v. Wade in 2022, as well as other temporal factors on
infant health outcomes.

@@

ggplot() +
  # First layer: Fill with death_rate
  geom_polygon(
    data = death_full,
    aes(x = long, y = lat, group = group, fill = death_rate),
    color = "black"
  ) +
  scale_fill_viridis_c(
    name = "Infant Death Rate\n(per 1,000 live births)",
    option = "C",
    na.value = "grey50",
    breaks = c(10, 15, 20,25),  # Custom breaks for the color legend
  ) +
  theme_void() +
  labs(
    caption = "Grey indicates missing data\nSource: Center for Reproductive Rights\nDate as of December, 2022"
  ) +
  theme(legend.position = "bottom")
```

@fig-choromap is a choropleth map visualizes the average infant death
rate per 1,000 live births across U.S. states, with data categorized by
color intensity. States with higher infant death rates are shown in
lighter shades (yellow), while states with lower rates are represented
in darker shades (purple). States with missing data are indicated in
grey. The map highlights disparities in infant mortality rates, with
some states, particularly in the southeastern U.S., exhibiting notably
higher rates compared to others. States like Alabama and Mississippi
show the highest rates, while states such as California and New York
exhibit lower rates.

@@

@tbl-motherage displays the average infant death rate (per 1,000 live
births) by the age group of mothers. It shows a clear trend: younger
mothers, particularly those aged 15–19 and 20–24, have higher infant
death rates (9.0 and 10.0, respectively), while older mothers in the
25–29 and 30–34 age groups have lower rates (7.4 and 6.6, respectively).
This suggests that maternal age may be a factor influencing infant
mortality, with younger mothers experiencing higher risks.

```{r}
#| label: tbl-motherage
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
library(kableExtra)
analysis_data_state <- analysis_data|>
  group_by(age_of_mother)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Age of Mother`=age_of_mother)

@@

### Mother's Single Race

The self-identified race of the mother, recorded as a single race
category. In this dataset, the available categories are: Asian, Black or
African American, and White. This classification facilitates the
analysis of maternal and infant health outcomes across these racial
groups.

```{r}
#| label: tbl-motherrace
#| tbl-cap: Average Infant Death Rate (per 1,000 live births) of different age group of Mother. It shows that infant death rates decrease with increasing maternal age, with the highest rate (10.0) among mothers aged 20–24 and the lowest (6.6) among those aged 30–34.
#| echo: false
#| warning: false
#| message: false
analysis_data_state <- analysis_data|>
  group_by(mothers_single_race)|>
  summarize(`Average Infant Death Rate (per 1,000 live births)` = round(mean(death_rate, na.rm = TRUE), 1)) |>rename(`Mother's Single Race`=mothers_single_race)

@@

For our analysis, we use a Bayesian Difference-in-Differences (DID) model to quantify the impact of abortion injunctions on the infant death rate across states in the USA. This method compares changes in infant death rates over time between a treatment group (states where abortion injunctions were implemented) and a control group (states where abortion remains legal), providing a robust framework for estimating causal effects.

The model focuses on the infant death rate per 1,000 live births as the dependent variable and includes the following independent variables:\
	-	Treatment Group Indicator `abortion_illegal`: A binary variable indicating the group: 1 if abortion is illegal in the state. 0 if abortion remains legal.\
	-	Time Indicator `after_injunction`: A binary variable indicating the time period: 0 for the period before the injunction implemented. 1 for the period after the injunction implemented.
	-	Interaction Term`after_injunction` × `abortion_illegal`: The interaction term captures the additional change in the outcome (e.g., percentage of support, behavior, or health-related metrics) for regions where abortion is illegal after the injunction implemented, beyond any changes observed in regions where abortion remains legal.

To account for potential regional variations in infant death rates, we include state as a random effect in the model. This allows us to control for unobserved heterogeneity across states that might introduce bias. Additionally, we include age_of_mother, mothers_single_race, and year_of_death as control variables to adjust for demographic and temporal factors that could influence infant mortality rates, ensuring that the estimated effects are not confounded by these characteristics.

By isolating the impact of the abortion injunction through this interaction term, the model identifies changes in infant death rates that can be attributed to the policy intervention. Additional details and diagnostics supporting the model’s implementation are provided in [Appendix-@sec-model-details].

## Difference-In-Difference Approach (DID)

DID is a causal inference method that estimates the impact of an intervention by comparing changes in outcomes over time between a treatment group and a control group. By assuming parallel trends in the absence of the intervention, the DID framework isolates the intervention’s effect through the interaction between time periods (pre- and post-intervention) and group membership. This makes it a robust method for evaluating policy changes and other significant events.

@@

In our analysis, we employ two models. The first model includes only the predictor variables and the random effect for state. The second model builds on the first by incorporating additional control variables to account for potential confounding factors.

Define $y_i$ as as the infant death rate per 1,000 live births.
$$y_i|\mu_i,\sigma \sim \mbox{Normal}(\mu_i, \sigma)\\
\text{First Model}: \mu_i = \beta_0 + \beta_1\cdot\text{After Injunction} + \beta_2\cdot\text{Abortion Illegal}+ \beta_3\cdot\text{After Injunction}\cdot\text{Abortion Illegal}_i+\gamma_j\\
\text{Second Model}:\mu_i = \beta_0 + \beta_1\cdot\text{After Injunction}_i + \beta_2\cdot\text{Abortion Illegal}_i+ \beta_3\cdot\text{After Injunction}_i\cdot\text{Abortion Illegal}_i\\+\beta_4\cdot\text{Year of Death}_i+\beta_5\cdot\text{Age of Mother}_i+\beta_6\cdot\text{Mother's Single Race}_i+\gamma_j\\
\beta_0 \sim \mbox{Normal}(0, 2.5) \\
\beta_1 \sim \mbox{Normal}(0, 2.5) \\
\beta_2 \sim \mbox{Normal}(0, 2.5) \\
\beta_3 \sim \mbox{Normal}(0, 2.5) \\
\beta_4 \sim \mbox{Normal}(0, 2.5) \\
\beta_5 \sim \mbox{Normal}(0, 2.5) \\
\beta_6 \sim \mbox{Normal}(0, 2.5) \\
\gamma_j \sim \text{Normal}(0, \sigma_j^2)\\
\sigma \sim \mbox{Exponential}(1)$$

The models are run in R (R Core Team 2023) using the rstanarm package of @rstanarm. The default priors from `rstanarm` are used for both GLM Bayesian models. We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm` used for both Gaussian Bayesian models.

@@

The Bayesian framework enhances the DID approach by incorporating prior information and quantifying uncertainty in a probabilistic manner. Using Markov Chain Monte Carlo (MCMC) sampling, the model estimates the posterior distributions of the parameters, offering flexibility in handling complex relationships and uncertainties in the data.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| include: false

@@

# Appendix {.unnumbered}

@@

```{r}
plot(first_model, pars = "(Intercept)", prob = 0.95)
```

```{r}
prior_summary(second_model)
```

@@

```{r}
plot(
  first_model,
  "areas"
)
plot(
  second_model,
  "areas"
)
```

```{r}
pp_check(second_model) +
  theme_classic() +
  theme(legend.position = "bottom")
```
